---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# readepi

*readepi* provides functions for importing epidemiological data into __R__ from
common _health information systems_.

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/readepi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/readepi/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/readepi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/readepi?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept) 
<!-- badges: end -->


## Installation

You can install the development version of readepi from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("epiverse-trace/readepi")
library(readepi)
```

# Reading data from file or directory
The **read_epi()** function can read read data from various source including various file formats. It is based on the **rio** R package, hence can import data from all file format supported by that one. See [here](https://cran.r-project.org/web/packages/rio/vignettes/rio.html) for more details about the file formats allowed by `rio`.    
Additionally, it can read in data from a space-separated file or read specific or all files from a specified directory. When reading data from file or directory, the function expects the following arguments:    
`file.path`: the path to the file to be read. When several files need to be imported from a directory, this should be the path to that directory,      
`sep`: the separator between the columns in the file. This is only required for space-separated files,     
`format`: a string used to specify the file format. This is useful when a file does not have an extension, or has a file extension that does not match its actual type,      
`which`: which a string used to specify which objects should be extracted (e.g. the name of the excel sheet to import),     
`pattern`: when specified, only files with this suffix will be imported from the specified directory.    
The function will return either a data frame (if import from file) or a list of data frames (if several files were imported from a directory).    

## importing data from JSON file
```{r}
file = "/Users/karimmane/Documents/Karim/LSHTM/Data/test.json"
data = readepi(file.path = file)
```

## importing data from excel file with 2 sheets
here we are importing from the second excel sheet
```{r}
file = "/Users/karimmane/Documents/Karim/LSHTM/Data/For_correlation.xlsx"
data = readepi(file.path = file, which = "Sheet2")
```

## importing data from several files in a directory
```{r}
# reading all files in the given directory
dir.path = "/Users/karimmane/Documents/Karim/LSHTM/Data"
data = readepi(file.path = dir.path)

# reading only txt files
data = readepi(file.path = dir.path, pattern = "txt")
```

# Reading data from relational database management systems (RDBMS): HDSS, EMRS, REDCap
Research data are usually stored in either relational databases or NoSQL databases. At the MRCG, project data are stored in relational databases. The HDSS and EMRS host databases that run under MS SQL Server, while REDCap (that uses an EAV schema) run under a MySQL server.    
To import data from a HDSS or EMRS (MS SQL Server) into R, some dependencies need to be installed first.    

## installation of dependencies
If you are using a Unix-based system, you will need to install the MS ODBC driver that is compatible with the version of the target MS SQL server. For **SQL server 2019, version 15.0**, we installed **ODBC Driver 17 for SQL Server** on the mac OS. This is compatible with the MRCG test server `robin.mrc.gm`. Details about how to install a driver can be found [here](https://learn.microsoft.com/en-us/sql/connect/odbc/linux-mac/install-microsoft-odbc-driver-sql-server-macos?view=sql-server-ver15).   
Once installed, the list of drivers can be displayed in R using: `odbc::odbcListDrivers()`.   
It is important to view the data that is stored in the MS SQL server. I recommend to install a GUI such as `Azure Data Studio`.     

## importing data from RDBMS into R
For testing purpose, we were given access to the test server `robin.mrc.gm`. Users should be granted with read access to be able to pull data from the test server.     
The **read_epi()** function can read data from the above RDBMS. It expects the following arguments:     
`credentials.file`: the path to the file with the user-specific credential details for the projects of interest. This is a tab-delimited file with the following columns:     
* user_name: the user name,    
* password: the user password (for REDCap, this corresponds to the **token** that serves as password to the project),    
* host_name: the host name (for HDSS and EMRS) or the URI (for REDCap),    
* project_id: the project ID (for REDCap) or the name of the database (for HDSS and EMRS) you are access to,   
* comment: a summary description about the project or database of interest,     
* dbms: the name of the DBMS: 'redcap' or 'REDCap' when reading from REDCap, 'sqlserver' or 'SQLServer' when reading from MS SQL Server,    
* port: the port ID (used for MS SQL Servers only).    
`project.id` for relational DB, this is the name of the database that contains the table from which the data should be pulled. Otherwise, it is the project ID you were given access to. Note that this should be similar to the value of the **project_id** field in the credential file.  
`driver.name` the name of the MS driver (only for HDSS and EMRS). use `odbc::odbcListDrivers()` to display the list of installed drivers,    
`table.name`: the name of the target table (only for HDSS and EMRS),   
`records`: a vector or a comma-separated string of a subset of subject IDs. When specified, only the records that correspond to these subjects will be imported,    
`fields`: a vector or a comma-separated string of column names. If provided, only those columns will be imported,     
`id.position`: the column position of the variable that unique identifies the subjects. This should only be specified when the column with the subject IDs is not the first column. default is 1.    

The function returns a list with 2 data frames (data and metadata) when reading from REDCap. A data frame otherwise.     

### importing from REDCap
we were given access to the **Pats__Covid_19_Cohort_1_Screening**.    
```{r}
credentials.file = "/Users/karimmane/Documents/Karim/LSHTM/Data/.credentials_karim.ini"

# reading all fields and records the project
data = readepi(credentials.file, project.id="Pats__Covid_19_Cohort_1_Screening")
project.data = data$data
project.metadeta = data$metadata

# reading specified fields and all records from the project
fields = c("day_1_q_ran_id","redcap_event_name","day_1_q_1a","day_1_q_1b","day_1_q_1c","day_1_q_1","day_1_q_2","day_1_q_3","day_1_q_4","day_1_q_5")
data = readepi(credentials.file, project.id="Pats__Covid_19_Cohort_1_Screening", fields = fields)

# reading specified records and all fields from the project
records = c("C10001/3","C10002/1","C10003/7","C10004/5","C10005/9","C10006/8","C10007/6","C10008/4","C10009/2","C10010/1")
data = readepi(credentials.file, project.id="Pats__Covid_19_Cohort_1_Screening", records =  records)

# reading specified records and fields from the project
data = readepi(credentials.file, project.id="Pats__Covid_19_Cohort_1_Screening", records =  records, fields = fields)
```

### importing from HDSS or EMRS
we were given access to the **IBS_BHDSS** database.   
```{r}
# reading all fields and all records from the `dss_events` table
data = readepi(credentials.file, project.id="IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events")

# reading specified fields and all records from the table
data = readepi(credentials.file, project.id="IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", fields = fields)

# reading specified records and all fields from the table
data = readepi(credentials.file, project.id="IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", records = records)

# reading specified fields and records from the table
data = readepi(credentials.file, project.id="IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", records = records, fields = fields)
```





## Development

### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/epiverse-trace/readepi/pulls).

Contributors to the project include:

* Thibaut Jombart (author)

### Code of Conduct
Please note that the linelist project is released with a 
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
