---
title: "readepi-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{readepi-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(readepi)
```

# Reading data from file or directory

`readepi` provides functions for reading data from `common health information systems` as well as from various file types. When several files of different formats are stored in the same folder, the user can provide `readepi` with a specific pattern, allowing reading only the matching files.     

When reading data from file or directory, the function expects the following arguments:   
`file.path`: the path to the file of interest. When several files need to be imported from a directory, this should be the path to that directory,      
`sep`: the separator between the columns in the file. This is only required for space-separated files,     
`format`: a string used to specify the file format. This is useful when a file does not have an extension, or has a file extension that does not match its actual type,      
`which`: a string used to specify which objects should be extracted (e.g. the name of the excel sheet to import),     
`pattern`: when provided, only files that contain this pattern will be imported from the specified directory.    
The function will return a list of data frames (if several files were imported from a directory).    

## importing data from JSON file
Many database management systems (DBMS) can export data into `JSON` format that can be read into R using the `readepi()` function.      
```{r eval=FALSE}
file <- system.file("extdata", "test.json", package = "readepi")
data <- readepi(file.path = file)
```

## importing data from excel file 
`readepi()` can import data from MS excel file with a single sheet. The user must specify the names of the excel sheets with the data of interest when the files contains several sheets.      
```{r eval=FALSE}
# IMPORTING DATA FROM THE SECOND EXCEL SHEET
file <- system.file("extdata", "test.xlsx", package = "readepi")
data <- readepi(file.path = file, which = "Sheet2")

# IMPORTING DATA FROM THE FIRST AND SECOND EXCEL SHEETS
file <- system.file("extdata", "test.xlsx", package = "readepi")
data <- readepi(file.path = file, which = c("Sheet2", "Sheet1"))
```

## importing data from several files in a directory
`readepi()` can be used to read data from multiple files that are all stored in the same directory. When there are different file format in that directory, the user is expected to specify the file type of interest with the `pattern` argument.     
```{r eval=FALSE}
# READING ALL FILES IN THE GIVEN DIRECTOR
dir.path <- system.file("extdata", package = "readepi")
data <- readepi(file.path = dir.path)

# READING ONLY '.txt' FILES
data <- readepi(file.path = dir.path, pattern = ".txt")

# READING '.txt' and '.xlsx' FILES
data <- readepi(file.path = dir.path, pattern = c(".txt", ".xlsx"))
```

# importing from DBMS (database management systems)
Users should be granted with read access to be able to pull data from the RDBMS. 

## Reading data from relational database management systems (RDBMS): HDSS, EMRS, REDCap
Research data are usually stored in either relational databases or NoSQL databases. For instance, at the MRCG, project data are stored in relational databases. The HDSS and EMRS host databases that run under MS SQL Server, while REDCap (that uses an EAV schema) run under a MySQL server.    
To import data from a HDSS or EMRS (MS SQL Server) into R, some dependencies need to be installed first.    

### installation of dependencies
If you are using a Unix-based system, you will need to install the MS ODBC driver that is compatible with the version of the target MS SQL server. For **SQL server 2019, version 15.0**, we installed **ODBC Driver 17 for SQL Server** on the mac OS which is compatible with the test server. 
The `install_odbc_driver()` function can be used to installed the chosen driver on your system.     

```{r eval=FALSE}
install_odbc_driver(driver_version = 17, force_install = FALSE)
```
Note that when prompted to agree with the license terms, enter **YES** in capital letters.    

If after running this command, you still have issues related to the MS SQL driver installation, use the links below for more details about how to install the **odbc** package in R and odbc drivers:     
[the odbc github page](https://github.com/r-dbi/odbc#installation)      

[here](https://learn.microsoft.com/en-us/sql/connect/odbc/linux-mac/install-microsoft-odbc-driver-sql-server-macos?view=sql-server-ver15)     

Once installed, the list of drivers can be displayed in R using:       
```{r eval=FALSE}
odbc::odbcListDrivers()
```

It is important to view the data that is stored in the MS SQL server. I recommend to install a GUI such as `Azure Data Studio`.         

### needed argument for importing data from RDBMS into R
To be able to read data from RDBMS, the **read_epi()** expects the following arguments:     
`credentials.file`: the path to the file with the user-specific credential details for the projects of interest. This is a tab-delimited file with the following 7  columns:    

* user_name: the user name,    
* password: the user password (for REDCap, this corresponds to the **token** that serves as password to the project),    
* host_name: the host name (for HDSS and EMRS) or the URI (for REDCap),    
* project_id: the project ID (for REDCap) or the name of the database (for HDSS and EMRS) you are access to,   
* comment: a summary description about the project or database of interest,     
* dbms: the name of the DBMS: 'redcap' or 'REDCap' when reading from REDCap, 'sqlserver' or 'SQLServer' when reading from MS SQL Server,    
* port: the port ID (used for MS SQL Servers only).    
`project.id` for relational DB, this is the name of the database that contains the table from which the data should be pulled. Otherwise, it is the project ID you were given access to. Note that this should be similar to the value of the **project_id** field in the credential file.  
`driver.name` the name of the MS driver (only for HDSS and EMRS). use `odbc::odbcListDrivers()` to display the list of installed drivers,    
`table.name`: the name of the target table (only for HDSS and EMRS),   
`records`: a vector or a comma-separated string of a subset of subject IDs. When specified, only the records that correspond to these subjects will be imported,    
`fields`: a vector or a comma-separated string of column names. If provided, only those columns will be imported,     
`id.position`: the column position of the variable that unique identifies the subjects. This should only be specified when the column with the subject IDs is not the first column. default is 1.    

Use the `show_example_file()` as below to view the credentials file used as template in the **readepi** package.      
```{r eval=FALSE}
# DISPLAY THE STRUCTURE OF THE CREDENTIALS FILE
show_example_file()
```
All demos in the sections below rely on the credentials stored in the template file, that is declared as following:
```{r eval=FALSE}
# DEFINING THE CREDENTIALS FILE
# credentials.file <- system.file("extdata", "test.ini", package = "readepi")
credentials.file <- system.file("extdata", "test.ini", package = "readepi")
```

### list the names of all tables in the database
To displays the list of tables from the database of interest, use:  
```{r eval=FALSE}
show_tables(
  credentials.file = credentials.file,
  project.id = "TEST_READEPI", # this is the database name
  driver.name = "ODBC Driver 17 for SQL Server"
)
```

Note that in the above, the value for the `project.id` argument is the name of the database of interest.    

### importing the data    
```{r eval=FALSE}
# READING ALL FIELDS AND ALL RECORDS FROM ONE TABLE (`DSS_EVENTS`)
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = "covid"
)

# READING SPECIFIED FIELDS AND ALL RECORDS FROM ONE TABLE
fields <- "date,sex,age" # same as fields = "idate, sex, age" and fields = c("date","sex","age")
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = "covid",
  fields = fields
)

# READING SPECIFIED RECORDS AND ALL FIELDS FROM ONE TABLE
records <- "e38000191,e38000247,e38000020,e38000130,e38000122,e38000223" # same as records=c("e38000191","e38000247","e38000020","e38000130","e38000122","e38000223") and records = "e38000191, e38000247, e38000020, e38000130, e38000122, e38000223"
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = "covid",
  records = records,
  id.position = 5
)

# READING SPECIFIED FIELDS AND RECORDS ONE THE TABLE
fields <- "date,sex,age,ccg_code"
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = "covid",
  records = records,
  fields = fields,
  id.col.name = "ccg_code"
)

# READING DATA FROM SEVERAL TABLES
table.names <- "covid,ebola,iris"
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = table.names
)

# READING DATA FROM SEVERAL TABLES AND SUBSETTING FIELDS ACROSS TABLES
table.names <- "covid, ebola, iris"
fields <- c(
  "date,sex,age,ccg_code", "id,date_of_onset,district",
  "Sepal.Length,Species"
) # the first string in the field vector corresponds to the name of the columns of interest from the first table specified in the `table.names` argument and so on...
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = table.names,
  fields = fields
)

# READING DATA FROM SEVERAL TABLES AND SUBSETTING RECORDS ACROSS TABLES
records <- c(
  "e38000191,e38000247,e38000020,e38000130,e38000122,e38000223",
  "9375,881,4539",
  "setosa"
) # "note that first string in the records vector corresponds to the records of interest from the first table specified in the `table.name` argument and so on... when the id column is not the first column in a table, use the `id.position`"
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = table.names,
  records = records,
  id.position = c(5, 1, 5)
)

# READING DATA FROM SEVERAL TABLES AND SUBSETTING RECORDS AND FIELDS ACROSS TABLES
fields <- c(
  "date,sex,age,ccg_code", "id,date_of_onset,district",
  "Sepal.Length,Species"
)
records <- c(
  "e38000191,e38000247,e38000020,e38000130,e38000122,e38000223",
  "9375,881,4539",
  "setosa"
)
data <- readepi(credentials.file,
  project.id = "TEST_READEPI",
  driver.name = "ODBC Driver 17 for SQL Server",
  table.name = table.names,
  records = records,
  fields = fields,
  id.col.name = c("ccg_code", "id", "Species")
)
```

## importing from REDCap
To import data from REDCap, the user must call the `readepi` function with the following arguments:     

* the credentials file (required)
* the project ID (required)
* the list of the desired records (optional)
* the list of the desired columns (optional)

Both the data and its associated metadata will be will be returned after a successful import.    

```{r eval=FALSE}
# READING ALL FIELDS AND RECORDS FROM A REDCap PROJECT
data <- readepi(
  credentials.file = credentials.file,
  project.id = "Pats__Covid_19_Cohort_1_Screening"
)
project.data <- data$data
project.metadeta <- data$metadata

# READING SPECIFIC FIELDS AND ALL RECORDS FROM THE PROJECT
fields <- c("day_1_q_ran_id", "redcap_event_name", "day_1_q_1a", "day_1_q_1b", "day_1_q_1c", "day_1_q_1", "day_1_q_2", "day_1_q_3", "day_1_q_4", "day_1_q_5")
data <- readepi(credentials.file,
  project.id = "Pats__Covid_19_Cohort_1_Screening",
  fields = fields
)

# READING SPECIFIC RECORDS AND ALL FIELDS FROM THE PROJECT
records <- c("C10001/3", "C10002/1", "C10003/7", "C10004/5", "C10005/9", "C10006/8", "C10007/6", "C10008/4", "C10009/2", "C10010/1")
data <- readepi(credentials.file,
  project.id = "Pats__Covid_19_Cohort_1_Screening",
  records = records,
  id.col.name = "day_1_q_ran_id"
)

# READING SPECIFIC RECORDS AND FIELDS FROM THE PROJECT
data <- readepi(credentials.file,
  project.id = "Pats__Covid_19_Cohort_1_Screening",
  records = records,
  fields = fields
)
project.data <- data$data
project.metadeta <- data$metadata
```


## importing from DHIS2
To import data from **DHIS2**, the user must call the `readepi` function with the following arguments:     

* the credentials file (required)
* the project ID (required)
* the dataset identifier (required)
* the organisation unit identifier (required)    
* the data element group (optional)    
* the start date for the time span of the values to export (required)    
* the end date for the time span of the values to export (required)   
* id.col.name the column name with the records of interest (optional)    
* the list of the desired records (optional)     
* the list of the desired columns (optional)    

```{r eval=FALSE}
# GETTING THE DATA ELEMENT IDENTIFIERS AND NAMES
data.elements <- get_data_elements(
  base.url = "https://play.dhis2.org/dev/",
  username = "admin",
  password = "district"
)

# GETTING THE ORGANISATION UNIT IDENTIFIERS AND NAMES
organisation.units <- get_data_sets(
  base.url = "https://play.dhis2.org/dev/",
  username = "admin",
  password = "district"
)

# GETTING THE DATA ELEMENT GROUP IDENTIFIERS AND NAMES
data.element.groups <- get_organisation_units(
  base.url = "https://play.dhis2.org/dev/",
  username = "admin",
  password = "district"
)

# READING THE DATASET ID `pBOMPrpg1QX`
data <- readepi(
  credentials.file = credentials.file,
  project.id = "DHIS2_DEMO",
  dataset = "pBOMPrpg1QX",
  organisation.unit = "DiszpKrYNg8",
  data.element.group = NULL,
  start.date = "2014",
  end.date = "2023"
)

# READING DATA FROM 2 DATASETS `pBOMPrpg1QX`
data <- readepi(
  credentials.file = credentials.file,
  project.id = "DHIS2_DEMO",
  dataset = "pBOMPrpg1QX,BfMAe6Itzgt", # same as dataset=c("pBOMPrpg1QX","BfMAe6Itzgt")
  organisation.unit = "DiszpKrYNg8",
  data.element.group = NULL,
  start.date = "2014",
  end.date = "2023"
)

# READING SPECIFIC DATA ELEMENTS FROM THE DATASET ID `pBOMPrpg1QX`
data.elts <- c("FTRrcoaog83", "eY5ehpbEsB7", "Ix2HsbDMLea")
data <- readepi(
  credentials.file = credentials.file,
  project.id = "DHIS2_DEMO",
  dataset = "pBOMPrpg1QX",
  organisation.unit = "DiszpKrYNg8",
  data.element.group = NULL,
  start.date = "2014",
  end.date = "2023",
  records = data.elts,
  id.col.name = "dataElement"
)

test.data <- data$data
```



