---
title: "readepi-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{readepi-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(readepi)
```

# Reading data from file or directory
## importing data from JSON file
```{r eval=FALSE}
file <- system.file("extdata", "test.json", package = "readepi")
data <- readepi(file.path = file)
```

## importing data from excel file with 2 sheets
here we are importing from the second excel sheet
```{r eval=FALSE}
file <- system.file("extdata", "test.xlsx", package = "readepi")
data <- readepi(file.path = file, which = "Sheet2")
```

## importing data from several files in a directory
```{r eval=FALSE}
# reading all files in the given directory
dir.path <- "inst/extdata"
data <- readepi(file.path = dir.path)

# reading only txt files
data <- readepi(file.path = dir.path, pattern = "txt")
```

# importing from REDCap
we were given access to the **Pats__Covid_19_Cohort_1_Screening**.    
```{r eval=FALSE}
# display the structure of the credentials file
show_example_file()
credentials.file <- system.file("extdata", "test.ini", package = "readepi")

# reading all fields and records the project
data <- readepi(credentials.file, project.id = "Pats__Covid_19_Cohort_1_Screening")
project.data <- data$data
project.metadeta <- data$metadata

# reading specified fields and all records from the project
fields <- c("day_1_q_ran_id", "redcap_event_name", "day_1_q_1a", "day_1_q_1b", "day_1_q_1c", "day_1_q_1", "day_1_q_2", "day_1_q_3", "day_1_q_4", "day_1_q_5")
data <- readepi(credentials.file, project.id = "Pats__Covid_19_Cohort_1_Screening", fields = fields)

# reading specified records and all fields from the project
records <- c("C10001/3", "C10002/1", "C10003/7", "C10004/5", "C10005/9", "C10006/8", "C10007/6", "C10008/4", "C10009/2", "C10010/1")
data <- readepi(credentials.file, project.id = "Pats__Covid_19_Cohort_1_Screening", records = records)

# reading specified records and fields from the project
data <- readepi(credentials.file, project.id = "Pats__Covid_19_Cohort_1_Screening", records = records, fields = fields)
```

# importing from HDSS or EMRS
To show the list of all tables in that database:      
```{r eval=FALSE}
showTables(credentials.file = system.file("extdata", "test.ini", package = "readepi"), project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server")
```

data import:      
```{r eval=FALSE}
# reading all fields and all records from one table (`dss_events`)
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events")

# reading specified fields and all records from one table
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", fields = fields)

# reading specified records and all fields from one table
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", records = records)

# reading specified fields and records one the table
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = "dss_events", records = records, fields = fields)

# reading data from several tables
table.names <- "accounts,account_books,account_currencies" # can also be table.names = c("account"s,"account_books","account_currencies")
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = table.names)

# reading data from several tables and subsetting fields across tables
table.names <- "accounts,account_books,account_currencies" # can also be table.names = c("account"s,"account_books","account_currencies")
# note that first string in the field vector corresponds to names to be subset from the first table specified in the `table.name` argument
fields <- c("id,name,balance,created_by", "id,status,name", "id,name")
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = table.names, fields = fields)

# reading data from several tables and subsetting records across tables
# note that first string in the records vector corresponds to the subject ID to be subset from the first table specified in the `table.name` argument and so on... When the ID column is not the first column in a table, use the `id.position`
records <- c("3,8,13", "1,2,3", "1")
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = table.names, records = records)

# reading data from several tables and subsetting records and fields across tables
fields <- c("id,name,balance,created_by", "id,status,name", "id,name")
records <- c("3,8,13", "1,2,3", "1")
data <- readepi(credentials.file, project.id = "IBS_BHDSS", driver.name = "ODBC Driver 17 for SQL Server", table.name = table.names, records = records, fields = fields)
```
